---
title: "Lab 4 - Nearest Neighbor Joining and Intro to Phylogeny"
author: "Becca DeCamp"
date: "7/9/2018"
output: html_document
---

## Introduction to Nearest Neighbor Joining and Phylogenetics
<br>

## Step 1: Getting Started
<br>
Log in to the SCC launch RStudio, as per ususal: 
<br>
```{bash, eval = F, echo = T}
#login
ssh -Y username@scc1.bu.edu
```
<br>
```{bash, eval = F, echo = T}
#access your directory and load your R space
cd /project/anth333/yourfilename
module load gcc
module load R
rstudio
```
<br>
Now, we will prepare our data for analysis. The main package we will be using is called APE (Analyses of Phylogenetics and Evolution), which uses a file format called a DNAbin. We will load in our data with the package vcfR, like we did in Lab 2, then we will convert our vcfR object to a DNAbin. 

```{r, results='hide', message=F, warning=F}
#First, load the packages: 

library(vcfR)
library(ape)
```
 (We will also install the package 'ggtree' which will help us graph our trees.)
```{r, eval = F, echo = T}
#(Now, we'll install ggtree from a site called Bioconductor:)

source("https://bioconductor.org/biocLite.R")
biocLite("ggtree")


library(ggtree)
```
<br>
Now, let's prepare our data. Again, I will be using the population YRI as an example. Remember to name your data according to your own population!
<br>
```{r}
#read in data
YRI <- read.vcfR("YRI.vcf")

#create DNAbin
YRIdna <- vcfR2DNAbin(YRI)
YRIdna
```

## Step 2: Creating A Distance Matrix Using Kimura's Evolutionary Model
<br>
As we learned in the introduction, Kimura's Neutral Theory is a popular model of molecular evolution. The function we will be using to create a distance matrix will 


